Double White Dwarf Mergers with CASTRO\\ I. Methodology and Code Verification

### 2015-02-23 papers/wdmerger_I/paper.tex
Our old strategy for tracking the stellar centers of mass failed when the stars got significantly far away from their expected circular orbit locations. This has been fixed by storing the center of mass at every time step and updating it with the old COM velocity. We then refine our guess using the location-weighted sums over the domain. We also make the cutoff more physically meaningful by using the approximation to the Roche radius of Eggleton (1983). To ensure continuity at a restart, I restored Mike's old solution of writing out the mass and COM to a file in the checkpoint, and reading it

@@

We track a number of diagnostic quantities at the end of every coarse grid timestep. 
For all simulations, we record the total energy (including the breakdown into
its components: kinetic, internal, gravitational potential, rotation; we note
that for the diagnostics we actually use $(\rho E)$ for calculation of the total energy,
rather than explicitly calculating the sum of kinetic and internal, as this is
the quantity that should be explicitly conserved), 
the total angular momentum, and the center of mass of the system. 
For stellar calculations, we separately record diagnostic 
information about the stars. Our strategy for tracking their 
locations is as follows: at the beginning of the calculation, we store the 
physical center of mass $\mathbf{x}_{\text{COM}}$ of the stars as determined 
by Kepler's third law. We also store the space velocity $\mathbf{v}_{\text{COM}}$ 
of the stars. Then, at each new time step we make a preliminary guess for their 
location by updating the location using the old velocity, 
$\mathbf{x}_{\text{COM}} \rightarrow \mathbf{x}_{\text{COM}} + \mathbf{v}_{\text{COM}} \Delta t$.
We then refine our guess for the location and velocity of each star by computing a
location-weighted sum of the mass and velocity over the computational domain. 
To do this, we need a cutoff for determining what counts as part of the primary 
and what counts as part of the secondary. For this we can use the formula provided by 
\citet{eggleton:1983} for the effective Roche radius $r_L$ of a star in a binary,
\begin{equation}
  \frac{r_L}{a} = \frac{0.49 q^{2/3}}{0.6^{2/3} + \text{ln}(1 + q^{1/3})}.
\end{equation}
In this formula we can use $q = M_S / M_P$ for obtaining the Roche radius of the secondary,
and use the inverse value of $q$ to obtain the Roche radius of the primary.
To a reasonable approximation, all material within the effective Roche radius
of each star can be counted as belonging to that star. From this we obtain the mass
of each star as well as its location and velocity.

@@

It is often stated in the literature that Eulerian methods for
hydrodynamics with grids fixed in space do not obey the Galilean
invariance of the underlying Euler equations, so that simulations
moving at a uniform bulk velocity will appear different than an
equivalent stationary simulation. If true, this means that codes such
as CASTRO may be fundamentally inappropriate for merger problems
(though any errors should be greatly diminished in a reference frame
that rotates with the initial binary orbit). Recently, this has come
up in two ways which are of note for us in the present study. We will
explain these situations and run tests to determine whether this
actually is a significant concern.

\citet{arepo} performed a Kelvin-Helmholtz instability test and showed
that (at low resolution) a fixed-grid code failed to develop the
expected fluid instability when the whole fluid was moving at a
strongly supersonic uniform velocity. This contrasted with the results
of the moving-mesh code AREPO being presented in that study, which
demonstrated Galilean invariance even at large bulk velocities. If
verified, this claim would have important consequences on how much we
can trust the ability of CASTRO to test the violent merger progenitor
model. Shearing between the material flowing out of the secondary and
material near the surface of the primary may trigger fluid
instabilities that play an important role in the evolution of that
gas, which is the site of the initial detonation in the prompt
explosion model. \citet{guillochon:2010} showed for their simulation
that Kelvin-Helmholtz instabilities produced this way may raise the
temperature of the accreting material enough to ignite a
detonation. Therefore if we are not correctly reproducing the
characteristics of the Kelvin-Helmholtz instability in the case where
there is significant mass motion on the grid, we cannot be confident
that a detonation (or lack thereof) is not numerically
seeded. However, \citet{robertson:2010} observe that Galilean
invariance of simulation results for the Euler equations occurs only
because of truncation error in the discretization of the fluid
equations. This takes the form of a numerical diffusion term which is
dependent on velocity (and also resolution). The advantage of a
moving-mesh code is that the mesh everywhere moves with the local flow
velocity, which substantially reduces the numerical
diffusion. \citeauthor{robertson:2010} argue that the differences seen
between the moving-mesh and fixed-grid code are caused by the
interaction of this numerical diffusion with small-scale instabilities
(which may be physical or numerical) which couple with and
fundamentally alter the large-scale modes. Small-scale instabilities
are seeded in this problem by the choice of a sharp initial
discontinuity between the fluids. Crucially though,
\citeauthor{robertson:2010} point out that this problem does not
converge with resolution and so it is not possible to know the correct
behavior of this problem. As such, we do not know whether the
small-scale modes found in AREPO are real, so the problem is not
useful in formally discriminating between methodologies. They instead
propose an alternate test with a smoother initial contact. This
converges to the same solution qualitatively in both the stationary
and bulk velocity cases, indicating that the code does generally
maintain Galilean invariance (to some specified error that depends on
resolution and the uniform flow speed).  We will see whether we can
reproduce this result, as well as a result by \cite{wadsley:2008}, who
used the FLASH code to simulate a hot bubble subject to mixing by the
Kelvin-Helmholtz instability, and also found that the mixing was
affected by a uniform bulk velocity.

@@

Following \cite{robertson:2010}, we set up a Kelvin-Helmholtz test in
the following way. The problem domain runs from 0 to 1 in both the $x$
and $y$ directions. Though this is a two-dimensional test, we run
CASTRO in 3D and replicate the problem along 8 cells in the $z$
direction; the physical size of these zones is determined so that $dz
= dx = dy$, a requirement at present in CASTRO. The problem involves a
fluid slab of density $\rho_1 = 2.0$ traveling rightward in the
$x$-direction at velocity $v_1 = 0.5$, sandwiched by a fluid of
density $\rho_2 = 1.0$ traveling leftward at velocity $v_2 =
-0.5$. The density gradient is in the $y$ direction, so this creates a
velocity shear along the interface between the fluids. The density and
velocity distribution on the computational domain are given by:
\begin{align}
  \rho &= \rho_1 + R(y)\left[\rho_2 - \rho_1\right] \\
  v_x  &= v_1 + R(y)\left[v_2 - v_1\right] \\
  v_y  &= v_{\text{bulk}} + v^\prime
\end{align}
Here $R(y)$ is a ramp function that describes the transition between
the two fluids, while $v_{\text{bulk}}$ is the bulk motion of the
fluid in the $y$ direction and $v^\prime$ is the velocity perturbation
that seeds the instability. The problem will be established for two
sets of initial conditions (ICs), which we follow
\citeauthor{robertson:2010} in calling ICs A and B. They differ in
their ramp function ($R_A$ and $R_B$ respectively), as well as the
initial perturbation ($v^\prime_A$ and $v^\prime_B$ respectively), and
the frequency of the perturbation ($n_A = 4$ and $n_B = 2$):
\begin{align}
  R_A &= \begin{cases} 0 & |y - 0.5| > 0.25 \\ 1 & |y - 0.5| < 0.25 \end{cases} \\
  R_B &= \Big\{\left[1 + e^{-2(y-0.25)/\Delta_y}\right]\left[1 + e^{2(y-0.75)/\Delta_y}\right]\Big\}^{-1} \\
  v^\prime_A &= w_0\, \text{sin}\left(n_A\, \pi\, x\right) \left\{e^{-(y-0.25)^2 / 2\sigma^2} + e^{-(y-0.75)^2/2\sigma^2}\right\} \\
  v^\prime_B &= w_0\, \text{sin}\left(n_B\, \pi\, x\right).
\end{align}
Here $w_0 = 0.1$ is the scale of the velocity perturbation, $\sigma =
0.05/\sqrt{2}$ controls the width of the Gaussian for IC A, and
$\Delta_y$ is the transition distance scale for the smooth ramp of IC
B. The pressure everywhere is set to $p = 2.5$, and we run this with a
gamma-law equation of state set to $\gamma = 5/3$. Plotfiles are
generated every 0.01 seconds, and the problem is run until $t = 2$.

We run the problem for $v_\text{bulk} = [0, 1, 3, 10, 30, 100]$, and
for each set of initial conditions run the problem at resolutions of
$64^2$, $128^2$, $256^2$, $512^2$. In addition, for each initial
condition we run a simulation at the very high resolution of $2560^2$
for the stationary problem only. This serves as a reference solution
to gauge the extent to which the bulk flow affects the development of
the fluid instability.

\subsubsection{Hot Bubble}\label{sec:hot_bubble}

>>>>

% allows us to force the location of a figure
\usepackage{float}

@@

We track a number of diagnostic quantities at the end of every coarse grid timestep. 
For all simulations, we record the total energy (including the breakdown into
its components: kinetic, internal, gravitational potential, rotation; we note
that for the diagnostics we actually use $(\rho E)$ for calculation of the total energy,
rather than explicitly calculating the sum of kinetic and internal, as this is
the quantity that should be explicitly conserved), 
the total angular momentum, and the center of mass of the system. 
For stellar calculations, we separately record diagnostic 
information about the stars. Our strategy for tracking their 
locations is as follows: at the beginning of the calculation, we store the 
physical center of mass $\mathbf{x}_{\text{COM}}$ of the stars as determined 
by Kepler's third law. We also store the space velocity $\mathbf{v}_{\text{COM}}$ 
of the stars. Then, at each new time step we make a preliminary guess for their 
location by updating the location using the old velocity, 
$\mathbf{x}_{\text{COM}} \rightarrow \mathbf{x}_{\text{COM}} + \mathbf{v}_{\text{COM}} \Delta t$.
We then refine our guess for the location and velocity of each star by computing a
location-weighted sum of the mass and velocity over the computational domain. 
To do this, we need a cutoff for determining what counts as part of the primary 
and what counts as part of the secondary. For this we can use the formula provided by 
\citet{eggleton:1983} for the effective Roche radius $r_L$ of a star in a binary,
\begin{equation}
  \frac{r_L}{a} = \frac{0.49 q^{2/3}}{0.6^{2/3} + \text{ln}(1 + q^{1/3})}.
\end{equation}
In this formula we can use $q = M_S / M_P$ for obtaining the Roche radius of the secondary,
and use the inverse value of $q$ to obtain the Roche radius of the primary.
To a reasonable approximation, all material within the effective Roche radius
of each star can be counted as belonging to that star. From this we obtain the mass
of each star as well as its location and velocity. Once we have the new centers of mass,
we compute the effective radius of each star at various density cutoffs. This involves 
computing the volume $V$ of all zones within the Roche radius of the star's center 
whose density is greater than the cutoff. We then compute $r_{\text{eff}} = (3V/4\pi)^{1/3}.$

@@

It is often stated in the literature that Eulerian methods for
hydrodynamics with grids fixed in space do not obey the Galilean
invariance of the underlying Euler equations, so that simulations
moving at a uniform bulk velocity will appear different than an
equivalent stationary simulation. If true, we need to understand 
the importance of this effect when deciding whether to trust the 
output of a code like CASTRO when applied for merger problems.
Recently, this has come up in two ways which are of note for us 
in the present study. We will explain these situations and display 
the results of tests we have run to determine whether this 
actually is a significant concern for our study.

\citet{arepo} (hereafter, S10) performed a Kelvin-Helmholtz instability test and showed
that (at low resolution) a fixed-grid code failed to develop the
expected fluid instability when the whole fluid was moving at a
strongly supersonic uniform velocity. (See also \citet{wadsley:2008}, 
who used the FLASH code to simulate a hot bubble subject to mixing 
by the Kelvin-Helmholtz instability, and also found that the mixing was affected by a 
uniform bulk velocity.) This contrasted with the results
of the moving-mesh code AREPO being presented in that study, which
demonstrated Galilean invariance even at large bulk velocities. 
Inability to correctly model the Kelvin-Helmholtz instability would 
ave important consequences for how much we can trust the ability of 
CASTRO to test the violent merger progenitor model. Shearing between 
the material flowing out of the secondary and material near the 
surface of the primary may trigger fluid
instabilities that play an important role in the evolution of that
gas, which is the site of the initial detonation in the prompt
explosion model. \citet{guillochon:2010} showed for their simulation
that Kelvin-Helmholtz instabilities produced this way may raise the
temperature of the accreting material enough to ignite a
detonation. Therefore if we are not correctly reproducing the
characteristics of the Kelvin-Helmholtz instability in the case where
there is significant mass motion on the grid, we cannot be confident
that a detonation (or lack thereof) is not numerically
seeded.

\begin{figure*}[ht]
  % Note that since we have periods in the filename, we need to double-
  % bracket the filename so that includegraphics doesn't think it's 
  % looking for something with the wrong extension. See:
  % http://tex.stackexchange.com/a/10575
  \centering
  \includegraphics[trim=0.75in 0.45in 0.85in 0.5in, clip, scale = 0.75]{{{plots/kh_t2.0_p2_low_res_collage}}}
  \caption{2D Kelvin-Helmholtz instability test at $t = 2.0$ for the initial 
    conditions given by Equations \ref{eq:kh_ic_b_ramp} and \ref{eq:kh_ic_b}. 
    The rows each represent a different bulk fluid velocity $v$ and 
    the columns each represent a grid resolution $n$ (the number of 
    zones per spatial dimension). The highest velocity simulation, 
    $v=100$, corresponds to approximately Mach 70. Compare to 
    \citet{robertson:2010}, Figure 7. \label{fig:kh_p2}}
\end{figure*}

\citet{robertson:2010} (hereafter, R10) observe that Galilean
invariance of simulation results for the Euler equations occurs only
because of truncation error in the discretization of the fluid
equations. This takes the form of a numerical diffusion term which is
dependent on velocity (and also resolution). The advantage of a
moving-mesh code is that the mesh everywhere moves with the local flow
velocity, which substantially reduces the numerical
diffusion. R10 argue that the differences seen
between the moving-mesh and fixed-grid code are caused by the
interaction of this numerical diffusion with small-scale instabilities
(which may be physical or numerical) which couple with and
fundamentally alter the large-scale modes. Small-scale instabilities
are seeded by the choice of a sharp initial discontinuity between the 
fluids in the problem posed by S10. Crucially though,
R10 point out that this problem does not
converge with resolution and so it is not possible to know the correct
behavior of this problem. As such, we do not know whether the
small-scale modes found in AREPO are real, and the problem is not
useful in formally discriminating between methodologies. They instead
propose an alternate test with a smoother initial contact. This
converges to the same solution qualitatively in both the stationary
and bulk velocity cases, indicating that the code does generally
maintain Galilean invariance (to some specified error that depends on
resolution and the uniform flow speed).  We will see whether we can
reproduce this result.

@@

Following \cite{robertson:2010}, we set up a Kelvin-Helmholtz test in
the following way. The problem domain runs from 0 to 1 in both the $x$
and $y$ directions. This is a two-dimensional test, so we run
CASTRO in 2D mainly to avoid extra computational expense; in 3D, it 
would merely involve replicating the problem in the $z$ direction.
The problem involves a fluid slab of density $\rho_2 = 2.0$ traveling rightward in the
$x$-direction at velocity $v_2 = 0.5$, sandwiched by a fluid of
density $\rho_1 = 1.0$ traveling leftward at velocity $v_1 =
-0.5$. The density gradient is in the $y$ direction, so this creates a
velocity shear along the interface between the fluids. The density and
velocity distribution on the computational domain are given by:
\begin{align}
  \rho &= \rho_1 + R(y)\left[\rho_2 - \rho_1\right] \\
  v_x  &= v_1 + R(y)\left[v_2 - v_1\right] \\
  v_y  &= v_{\text{bulk}} + v^\prime
\end{align}
Here $R(y)$ is a ramp function that describes the transition between
the two fluids, while $v_{\text{bulk}}$ is the bulk motion of the
fluid in the $y$ direction and $v^\prime$ is the velocity perturbation
that seeds the instability. The problem will be established for two
sets of initial conditions (ICs), which we follow
R10 in calling ICs A and B. They differ in
their ramp function ($R_A$ and $R_B$ respectively), as well as the
initial perturbation ($v^\prime_A$ and $v^\prime_B$ respectively), and
the frequency of the perturbation ($n_A = 4$ and $n_B = 2$):
\begin{align}
  R_A &= \begin{cases} 0 & |y - 0.5| > 0.25 \\ 1 & |y - 0.5| < 0.25 \end{cases} \label{eq:kh_ic_a_ramp}\\
  R_B &= \Big\{\left[1 + e^{-2(y-0.25)/\Delta_y}\right]\left[1 + e^{2(y-0.75)/\Delta_y}\right]\Big\}^{-1} \label{eq:kh_ic_b_ramp}\\
  v^\prime_A &= w_0\, \text{sin}\left(n_A\, \pi\, x\right) \left\{e^{-(y-0.25)^2 / 2\sigma^2} + e^{-(y-0.75)^2/2\sigma^2}\right\} \label{eq:kh_ic_a}\\
  v^\prime_B &= w_0\, \text{sin}\left(n_B\, \pi\, x\right). \label{eq:kh_ic_b}
\end{align}

\begin{figure*}[ht]
  \centering
  \includegraphics[trim=0.75in 0.45in 0.85in 0.5in, clip, scale = 0.75]{{{plots/kh_t2.0_p3_low_res_collage}}}
  \caption{2D Kelvin-Helmholtz instability test at $t = 2.0$ for the initial 
    conditions given by Equations \ref{eq:kh_p3_rho} through \ref{eq:kh_p3_vy}. 
    The meaning of the rows and columns is the same as in Figure \ref{fig:kh_p2}.
    \label{fig:kh_p3}}
\end{figure*}

Here $w_0 = 0.1$ is the scale of the velocity perturbation, $\sigma =
0.05/\sqrt{2}$ controls the width of the Gaussian for IC A, and
$\Delta y = 0.05$ is the transition distance scale for the smooth ramp of IC
B. The pressure everywhere is set to $p = 2.5$, and we run this with a
gamma-law equation of state set to $\gamma = 5/3$. Plotfiles are
generated every 0.05 seconds, and the problem is run until $t = 2$.

We run the problem for $v_\text{bulk} = [0, 1, 3, 10, 30, 100]$, and
for each set of initial conditions run the problem at resolutions of
$64^2$, $128^2$, $256^2$, $512^2$. For context, in these units the 
sound speed is $c\approx 0.7$. In addition, for each initial
condition we run simulations at the higher resolutions of $1024^2$,
$2048^2$, and $4096^2$ for the stationary problem only. These serve 
as a reference solution to gauge the extent to which the bulk flow 
affects the development of the fluid instability, and to determine 
if the problem is numerically converged.

We find the same result as R10 for IC A, which is equivalent to the 
test proposed by S10: at low resolutions and high bulk velocity, the 
Kelvin-Helmholtz instability completely fails to develop. However, 
the problem does not converge even qualitatively at the highest 
resolutions we used. Our results are very similar to Figure 3 of 
R10 so we do not show them here. For IC B, our results can be seen 
for the normal resolutions and all velocities in Figure \ref{fig:kh_p2}.
At low resolutions and very large bulk velocities, the fluid 
does get significantly disrupted by numerical error. However, this 
effect quickly converges away with resolution and qualitatively 
at $512^2$ resolution the solution is nearly identical to the stationary 
$v=0$ problem. We agree with R10 that this problem does converge 
with resolution and is not subject to numerically-seeded secondary 
instabilities at the stopping time. This is evident even at low resolutions
by examining the first row of Figure \ref{fig:kh_p2}.

\begin{figure*}[ht]
  \includegraphics[trim = 0in 1.25in 0.25in 1.25in, clip, scale=0.65]{{{plots/kh_p3_high_res_collage}}}
  \caption{Time series of the Kelvin-Helmholtz problem proposed by \citet{mcnally:2012}
    as the simulation is just starting to go non-linear. The rows represent resolution, 
    where $n$ is the number of grid cells per spatial dimension, and the columns are 
    different snapshots in time.\label{fig:kh_p3_high_res}}
\end{figure*}

\citet{mcnally:2012} published another Kelvin-Helmholtz problem that 
is well-posed in the sense that it converges with resolution and 
is not subject to uncontrollable numerical instabilities. Though they 
were not explicitly interested in the question of Galilean invariance, 
we visit that issue here to see what can be learned. The initial 
conditions for this problem are:
\begin{align}
  \rho &= \begin{cases} \rho_1 - \rho_m e^{\frac{y-0.25}{\Delta y}} & 0.25 > y \geq 0 \\ 
                        \rho_2 + \rho_m e^{\frac{0.25-y}{\Delta y}} & 0.5 > y \geq 0.25 \\
                        \rho_2 + \rho_m e^{\frac{y-0.75}{\Delta y}} & 0.75 > y \geq 0.5 \\
                        \rho_1 - \rho_m e^{\frac{0.75-y}{\Delta y}} & 1 > y \geq 0.75 \end{cases} \label{eq:kh_p3_rho}\\
  v_x &= \begin{cases} v_1 - v_m e^{\frac{y-0.25}{\Delta y}} & 0.25 > y \geq 0 \\
                       v_2 + v_m e^{\frac{0.25-y}{\Delta y}} & 0.5 > y \geq 0.25 \\
                       v_2 + v_m e^{\frac{y-0.75}{\Delta y}} & 0.75 > y \geq 0.5 \\
                       v_1 - v_m e^{\frac{0.75-y}{\Delta y}} & 1 > y \geq 0.75 \end{cases} \label{eq:kh_p3_vx} \\
  v_y &= w_0\, \text{sin}\left(4\pi x\right). \label{eq:kh_p3_vy}
\end{align}
Here $\Delta y = 0.025$, $w_0 = 0.01$, $v_m = (v_1 - v_2) / 2$, $\rho_m = (\rho_1 - \rho_2) / 2$, 
and the other symbols have the same meaning as above (this means the flow direction
is reversed compared to the original paper, so as to achieve consistency with the 
other simulations presented here). We run this problem at all the same resolutions 
and bulk velocities as the previous two problems. The results for the normal resolutions 
at $t = 2.0$ are displayed in Figure \ref{fig:kh_p3}. We see a similar pattern as for 
the test proposed by R10: as we get to higher flow speeds we need to have higher 
spatial resolution to compensate for the increased numerical diffusion. However, 
the accuracy is much lower for the highest bulk velocities. This is probably because 
the initial conditions feature a transition that is twice as sharp (compare the $\Delta y$ 
parameters in each case). This means that there are smaller scale numerical 
perturbations on the grid, which can be amplified by numerical noise in the time 
evolution of the system.

\citet{hopkins:2014} performed this test as part of the testing of their code GIZMO. 
They showed the late-time evolution of this system, when non-linear effects have 
taken over and significantly disrupted the initial flow. However, at low resolution the 
tested grid algorithm had failed to disrupt both for $v = 0$ and $v = 10$. We too ran 
this problem until $t = 10$, and confirm that the Kelvin-Helmholtz instability damps out 
at low resolution but goes strongly non-linear and disrupts the flow at high resolution.
However, we strongly emphasize the point that this does not objectively 
demonstrate a defiency in fixed-grid codes for this problem. We can only determine 
the validity of a method when we have a trustworthy, converged solution to compare to, and 
this is lacking for this problem at late times. As observed by \citeauthor{mcnally:2012}, 
this is because the secondary instabilities form for this problem when the whorls of 
the Kelvin-Helmholtz tendrils stretch out and create gradients that approach the 
grid resolution. This is prime breeding ground for numerical noise. But because 
the nature of this noise will depend on the resolution, it will be very different for 
simulations at different resolutions. If these instabilities are seeded because of 
this resolution-dependent noise and are not seeded instead in a controlled manner 
such that they appear at the same time and location, then we simply cannot draw 
any conclusions that bear on the question of verification from this test at late times. 
Figure \ref{fig:kh_p3_high_res} provides a sense of this by examining the crucial 
time at which the transition from the linear to the non-linear regime is occurring. 
At all of these very high resolutions the secondary instabilities develop, but they 
occur at different times and have different spatial scales for each resolution.

We conclude that large bulk motions of fluid can have very significant effects 
on numerical calculations of shear mixing in fixed-grid codes, but that this effect 
diminishes with increasing resolution. As a result, we need only be confident that we are 
sufficiently resolving the major mixing regions on the white dwarf surfaces. 
If we find that this mixing occurs near the grid resolution scale, this will imply 
that we need to ramp up the resolution in these regions using AMR. If this becomes 
too expensive, we would need to be skeptical of any conclusions that could be drawn 
about the effect of the mixing on the nuclear burning.
